Stochastic Newton with Arbitrary Sampling

### 2023-02-25 paper/main.tex
Update main.tex

	The problem of minimizing the average of a large number of sufficiently smooth and strongly convex functions is ubiquitous in machine learning. Stochastic first-order methods for this problem of Stochastic Gradient Descent type are well studied. In turn, second-order methods, such as Newton, have certain advances since they can adapt to the curvature of the problem. They are also known for their fast convergence rates. But stochastic variants of Newton-type methods are not studied as good as SGD-type ones and have limitations on the batch size. Dmitry Kovalev et al proposed a method which requires no limitations on batch sizes. Our goal is to explore this method with different sampling strategies that lead to practical improvements.

@@

	The stochastic approach is used, because $n$ is very large and it is computationally difficult to evaluate the gradient of each $f_i$ on each step. The Stochastic Gradient Descent (SGD) method \cite{SGD-1} calculates the gradients of some randomly chosen $f_i$, which has cheap iterations compared to the deterministic algorithm. Such methods have two certain advantages. Firstly, they do not rely on the number of used $f_i$ on a single step, which is called the batch size. Therefore, these methods can be applied even with small batches. \\

	Secondly, there are ways to solve some common problem that arises due to the stochastic nature of the methods. The fact is that stochastic methods provide convergence to a neighbourhood of the solution only \cite{sgd-hogwild, sgd-general-analysis}. Its size is proportional to the variance of the stochastic gradient. So the second advantage of SGD is presence of so-called variance-reduced methods \cite{exp-convergence, advances-NIPS, unified-sgd, one-method}, which mitigate the mentioned effect. However, these methods' computational complexity depends on the curvature of the problem, which is called the condition number and is defined as the ratio of Lipschitzness and strong convexity parameters. \\

	This leads to usage of second-order methods, such as the Newton method \cite{Nesterov-introductory, Newton-convergence, RSN}. Taking into account the derivatives of the second order makes it possible to adjust the algorithm's step sizes to the curvature of the problem \cite{Nesterov-introductory}. Unfortunately, much less literature has been written on this topic than about first-order methods. Some proposed algorithms need extra assumptions \cite{Newton-convergence, RSN} or regularization \cite{cubic-regularization, randomized-cubic, uniform-convex-cubic} to converge. Some works \cite{proximal-quasi, lbfgs, stoch-bfgs} provide stochastic quasi-Newton methods, whose computational complexity exceeds such in the variance-reduced SGD variants \cite{advances-NIPS}. \\

	In addition, there are many algorithms presented \cite{sub-sampled, exact-inexact, variance-reduced-Newton, zhang2022adaptive, tripuraneni2018stochastic, zhou2020stochastic}, that need large batch sizes. Particularly, the required batch size is commonly quadratically proportional to the inverse of the desired accuracy. That means that one need to evaluate a large number of $f_i$ Hessians. And this decreases the profit gained by adding randomness into the algorithm, because these batch sizes can become as big as $n$. \\

	Dmitry Kovalev, Konstantin Mishchenko and Peter Richtarik proposed \cite{kovalev2019stochastic} a simple Stochastic Newton algorithm, which can work with small batches, even with batches of size one. Their algorithm does not provide unbiased estimates but nevertheless shows good convergence. This is achieved by developing new Lyapunov functions that are specific to second-order methods. Our goal is to apply different sampling strategies to this method and explore their performance. \\

	The basic set of sampling strategies can be taken from \cite{richtarik2016parallel}, where these strategies are applied to the Parallel Coordinate Descent methods. It is also known that so-called Importance Sampling can improve the SGD performance \cite{https://doi.org/10.48550/arxiv.1401.2753, 9413313}, since it can reduce the stochastic gradient variance. \\

>>>>

	The problem of minimizing the average of a large number of sufficiently smooth and strongly convex functions is ubiquitous in machine learning. Stochastic first-order methods for this problem of Stochastic Gradient Descent type are well studied. In turn, second-order methods, such as Newton, have certain advances since they can adapt to the curvature of the problem. They are also known for their fast convergence rates. But stochastic variants of Newton-type methods are not studied as good as SGD-type ones and have limitations on the batch size. \cite{kovalev2019stochastic} proposed a method which requires no limitations on batch sizes. Our goal is to explore this method with different sampling strategies that lead to practical improvements.

@@

	Typically, $n$ is very large for modern real-world problems. Thus, the stochastic approach is used because it is computationally difficult to evaluate the gradient of all $f_i$ at each step. The Stochastic Gradient Descent (SGD) method \cite{SGD-1} calculates the gradients of some randomly chosen $f_i$ which leads to cheaper computation per-iteration cost compared to the vanilla Gradient Descent (GD). The analysis of SGD and its modifications is rich and well explored. Firstly, the theory of SGD-type methods does not restrict the batch size. Therefore, these algorithms can be applied even with small batches. It is known that simple SGD converges only to a neighbourhood of the solution only \cite{sgd-hogwild, sgd-general-analysis} whose size is proportional to the variance of the stochastic gradient. However, there are techniques (aka variance reduction) to solve this issue. These techniques \cite{exp-convergence, advances-NIPS, unified-sgd, one-method} modify the update rule of vanilla SGD which allows to mitigate the aforementioned effect without changing per-iteration cost. However, the main disadvantage of all gradient-type methods is that a computation complexity depends on the curvature of the problem, which is called the condition number and is defined as the ratio of Lipschitzness and strong convexity parameters. \\

	This is a place where second-order methods such as the Newton method \cite{Nesterov-introductory, Newton-convergence, RSN} come to play. Taking into account second-order derivatives it possible to adjust the algorithm's step sizes to the curvature of the problem \cite{Nesterov-introductory}. Unfortunately, much less work has been done in the direction of stochastic Newton-type methods. Many algorithms \cite{sub-sampled, exact-inexact, variance-reduced-Newton, zhang2022adaptive, tripuraneni2018stochastic, zhou2020stochastic} require large batch sizes. In particular, the required batch size is commonly quadratically proportional to the inverse of the desired accuracy. That means that one need to evaluate a large number of $f_i$ Hessians which sometimes can be much larger than $n$. \cite{kovalev2019stochastic} proposes a simple Stochastic Newton algorithm, which can work with batches of any size. Their algorithms achieve local linear and super-linear convergence. \\

	In practice, various sampling strategies are used for SGD-type algorithms to improve further the performance. One of the most famous sampling mechanisms is so-called Importance Sampling \cite{https://doi.org/10.48550/arxiv.1401.2753, 9413313}. The idea is to compute gradients of the functions that have more impact on the problem. \cite{richtarik2016parallel} studies many other sampling mechanisms. We aim to analyse such strategies, but for Algorithms $1$ and $2$ of \cite{kovalev2019stochastic} to improve the theoretical and practical applications of algorithms. We investigate various sampling strategies supporting them with  rigorously constructed experiments.
